\mytitle

### 2021-12-20 paper.tex
runtime -> run-time

Many of the publicly available implementations, e.g.\ \cite{root,gsl},
use formulations for the approximation of the dilogarithm with a
strictly sequential execution of floating\-/point operations.  The
development of modern CPUs with support for instruction\-/level
parallelism (ILP) opens the opportunity for a further reduction of the
runtime of the calculation of the dilogarithm.  To make use of ILP,
the algorithm to calculate the dilogarithm must be formulated
appropriately in order to allow for the parallel execution of multiple
floating\-/point operations.  In this paper a method for the
calculation of the real dilogarithm $\Li$ is presented, which makes
use of ILP, potentially resulting in a reduced runtime compared to
many algorithms presented so far.

@@

As is well known, the performante of this naive implementation can be
improved by applying Horner's scheme, which reduces the number of
expensive floating\-/point multiplications to 7 and requires 6
additions:
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \bigg\{1 + x \bigg[-\frac{1}{2} + x \bigg(\frac{1}{3} + x \bigg\{-\frac{1}{4} \\
  &+ x \bigg[\frac{1}{5} + x\bigg(-\frac{1}{6} + \frac{x}{7}\bigg)\bigg]\bigg\}\bigg)\bigg]\bigg\}.
\end{split}\label{eq:horner}%
\end{align}
%
A C implementation using Horner's scheme could read
%
\lstinputlisting[language=C]{src/log1p_horner.c}
%
Note, that the formulation using Horner's scheme requires the
evaluation of the floating\-/point operations in strictly sequential
order.  However, the significant reduction of the number of
floating\-/point multiplications, compared to the naive
implementation, usually results in a shorter runtime.

There are several different possibilities to reduce the runtime of
Horner's method by rewriting the series expansion of $\ln(1+x)$ such
that more terms can be calculated in parallel, while the number
floating\-/point operations is still kept small.  One possibility is
to split the r.h.s.\ of Eq.~\eqref{eq:horner} into two distinct sums,
one containing the even powers and the other containing the odd powers
of $x$,
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \left\{1 + y \left[\frac{1}{3} + y
  \left(\frac{1}{5} + \frac{y}{7}\right)\right]\right\} \\
  &- y\left[\frac{1}{2} + y\left(\frac{1}{4} + \frac{y}{6}\right)\right],
\end{split}\label{eq:split}%
\end{align}
%
where $y=x^2$.  This formulation requires 8 floating\-/point
multiplications and 6 additions, i.e.\ one more multiplication than
Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can then
be calculated in parallel, leading to a potential speed-up if the cost
of the additional multiplication is smaller than the gain by the
parallel execution.  A C implementation could read:
%
\lstinputlisting[language=C]{src/log1p_split.c}
%
Another possibility is to rewrite the r.h.s.\ of Eq.~\eqref{eq:taylor}
using Estrin's scheme \cite{estrin},
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
    y \\ &+\left[\left(\frac{x}{7}-\frac{1}{6}\right)
    y+\frac{x}{5}-\frac{1}{4}\right] z,
\end{split}\label{eq:estrin}%
\end{align}
%
where $y=x^2$ and $z=y^2$.  This form also requires 8 floating\-/point
multiplications and 6 additions, as in Eq.~\eqref{eq:split}.  However,
in Estrin's form more terms can be executed in parallel, leading to a
potential further speed-up.  A C implementation using Estrin's scheme
could read:
%
\lstinputlisting[language=C]{src/log1p_estrin.c}

@@

\section{Runtime comparison}

In \tabref{tab:runtime} the runtime of different C implementations is
compared to the implementation given in the appednix on two different
CPU architectures.  On each architecture the implentations are
compiled with \texttt{g++} with optimization \texttt{-O2}.  The shown
runtime denotes the total wall clock time for the calculation of
$\Li(x)$ for $10^6$ uniformly distributed random values $x\in[0,1/2]$.
The choice of this interval ensures that the measurement is only
sensitive to the formulation of the approximant, but not to the
transformation onto the interval $[0,1/2]$.%
%
\begin{table}[tb]
  \centering
  \caption{Runtime in seconds for the calculation of $\Li(x)$ for
    different implementations on different CPU architectures with
    different compilers.}
  \begin{tabular}{lcc}
    \toprule
    \multirow{2}{*}{Implementation} & i7-4700MQ & i7-5600U \\
    & \texttt{g++} 9.3.0 & \texttt{g++} 10.2.1 \\
    \midrule
    Eq.~\eqref{eq:final} (this paper) & $0.0076$ & $0.0058$ \\
    Morris (DILOG 0011) \cite{morris} & $0.0115$ & $0.0085$ \\
    Cephes \cite{cephes}              & $0.0136$ & $0.0119$ \\
    Algorithm 327 \cite{koelbigDilog} & $0.0229$ & $0.0172$ \\
    Chebyshev \cite{luke,root}        & $0.0718$ & $0.0515$ \\
    Algorithm 490 \cite{ginsberg}     & $0.1057$ & $0.0576$ \\
    GSL \cite{gsl}                    & $0.2091$ & $0.0898$ \\
    \bottomrule
  \end{tabular}
  \label{tab:runtime}
\end{table}%

We find that the implementation presented in this paper and the
approach presented in Ref.~\cite{morris} (DILOG 0011) are fastest
among the ones shown in the table.  Both approaches use a rational
function approximant with the same number of terms.  However, the
approximant from Ref.~\cite{morris} has been implemented (for
comparison) to use Horner's scheme to evaluate the numerator and the
denominator.  Thus, the runtime difference of $\sim 30\%$ between
the two stems only from the fact that the implementation presented
in this paper uses Estrin's scheme to evaluate the numerator and the
denominator and as such can make use of ILP, while in the
implementation of Ref.~\cite{morris} the numerator and the denominator
must be evaluated sequentially.

The third fastest implementation is provided by Cephes \cite{cephes},
which implements a rational function approximation for Spence's
function $\operatorname{Sp}(x)=\Li(1-x)$.  The implementation provided
by Cephes is by a factor two slower than the implementation prestend
in this paper.  This longer runtime has several reasons: For one,
Cephes uses a rational function approximation with more terms in the
numerator and denominator.  Furthermore, Cephes evaluates the
numerator and denominator polynomials using Horner's scheme, which is
purely sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$
  twice, which leads to a further increased runtime when a benchmark
  interval with $x>1$ is chosen.}

Algorithm 327 \cite{koelbigDilog}, Algorithm 490 \cite{ginsberg} and
the implementation found in the GNU Scientific Library (GSL)
\cite{gsl} use series expansions, which are summed using Horner's
scheme.  These algorithms thus do not make use of ILP, which is one
reason for their longer runtime.  Furthermore, Algorithm 490 and most
series expansions in the GSL use floating\-/point divisions inside
loops, which may be another reason for the increased runtime.

A representation in terms of Chebyshev polynomials \cite{luke} is
implemented in the ROOT program library \cite{root}.  In this
implementation the polynomials are summed using Clenshaw's algorithm,
which is also purely sequencial and slightly more complex than
Horner's scheme.  This is a reason why the implementation in
\cite{root} has a larger runtime than the rational function
approximants discussed above.

@@

In this paper different approaches for the numerical calculation of
the real dilogarithm have been compared w.r.t.\ their potential for
the use of instruction\-/level (ILP) parallelism.  We found that a
rational minimax approximant in combination with Estrin's scheme to
evaluate the numerator/denominator polynomials provides both a
sufficient numerical precision and a high degree of ILP, potentially
leading to a reduced runtime.  An approximant and a corresponding C
implementation have been proposed and compared to several existing
implementations w.r.t.\ their runtime.  The C implementation of the
presented approximant can be found in the appendix.  C++ and FORTRAN
implementations can be found in the Polylogarithm library
\cite{polylogarithm}.  Implementations in Julia and Rust can be found
in \cite{PolyLog.jl} and \cite{polylog}, respectively.

>>>>

Many of the publicly available implementations, e.g.\ \cite{root,gsl},
use formulations for the approximation of the dilogarithm with a
strictly sequential execution of floating\-/point operations.  The
development of modern CPUs with support for instruction\-/level
parallelism (ILP) opens the opportunity for a further reduction of the
run\-/time of the calculation of the dilogarithm.  To make use of ILP,
the algorithm to calculate the dilogarithm must be formulated
appropriately in order to allow for the parallel execution of multiple
floating\-/point operations.  In this paper a method for the
calculation of the real dilogarithm $\Li$ is presented, which makes
use of ILP, potentially resulting in a reduced run\-/time compared to
many algorithms presented so far.

@@

As is well known, the performante of this naive implementation can be
improved by applying Horner's scheme, which reduces the number of
expensive floating\-/point multiplications to 7 and requires 6
additions:
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \bigg\{1 + x \bigg[-\frac{1}{2} + x \bigg(\frac{1}{3} + x \bigg\{-\frac{1}{4} \\
  &+ x \bigg[\frac{1}{5} + x\bigg(-\frac{1}{6} + \frac{x}{7}\bigg)\bigg]\bigg\}\bigg)\bigg]\bigg\}.
\end{split}\label{eq:horner}%
\end{align}
%
A C implementation using Horner's scheme could read
%
\lstinputlisting[language=C]{src/log1p_horner.c}
%
Note, that the formulation using Horner's scheme requires the
evaluation of the floating\-/point operations in strictly sequential
order.  However, the significant reduction of the number of
floating\-/point multiplications, compared to the naive
implementation, usually results in a shorter run\-/time.

There are several different possibilities to reduce the run\-/time of
Horner's method by rewriting the series expansion of $\ln(1+x)$ such
that more terms can be calculated in parallel, while the number
floating\-/point operations is still kept small.  One possibility is
to split the r.h.s.\ of Eq.~\eqref{eq:horner} into two distinct sums,
one containing the even powers and the other containing the odd powers
of $x$,
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \left\{1 + y \left[\frac{1}{3} + y
  \left(\frac{1}{5} + \frac{y}{7}\right)\right]\right\} \\
  &- y\left[\frac{1}{2} + y\left(\frac{1}{4} + \frac{y}{6}\right)\right],
\end{split}\label{eq:split}%
\end{align}
%
where $y=x^2$.  This formulation requires 8 floating\-/point
multiplications and 6 additions, i.e.\ one more multiplication than
Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can then
be calculated in parallel, leading to a potential speed-up if the cost
of the additional multiplication is smaller than the gain by the
parallel execution.  A C implementation could read:
%
\lstinputlisting[language=C]{src/log1p_split.c}
%
Another possibility is to rewrite the r.h.s.\ of Eq.~\eqref{eq:taylor}
using Estrin's scheme \cite{estrin},
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
    y \\ &+\left[\left(\frac{x}{7}-\frac{1}{6}\right)
    y+\frac{x}{5}-\frac{1}{4}\right] z,
\end{split}\label{eq:estrin}%
\end{align}
%
where $y=x^2$ and $z=y^2$.  This form also requires 8 floating\-/point
multiplications and 6 additions, as in Eq.~\eqref{eq:split}.  However,
in Estrin's form more terms can be executed in parallel, leading to a
potential further speed-up.  A C implementation using Estrin's scheme
could read:
%
\lstinputlisting[language=C]{src/log1p_estrin.c}

@@

\section{Run\-/time comparison}

In \tabref{tab:runtime} the run\-/time of different C implementations is
compared to the implementation given in the appednix on two different
CPU architectures.  On each architecture the implentations are
compiled with \texttt{g++} with optimization \texttt{-O2}.  The shown
run\-/time denotes the total wall clock time for the calculation of
$\Li(x)$ for $10^6$ uniformly distributed random values $x\in[0,1/2]$.
The choice of this interval ensures that the measurement is only
sensitive to the formulation of the approximant, but not to the
transformation onto the interval $[0,1/2]$.%
%
\begin{table}[tb]
  \centering
  \caption{Run\-/time in seconds for the calculation of $\Li(x)$ for
    different implementations on different CPU architectures with
    different compilers.}
  \begin{tabular}{lcc}
    \toprule
    \multirow{2}{*}{Implementation} & i7-4700MQ & i7-5600U \\
    & \texttt{g++} 9.3.0 & \texttt{g++} 10.2.1 \\
    \midrule
    Eq.~\eqref{eq:final} (this paper) & $0.0076$ & $0.0058$ \\
    Morris (DILOG 0011) \cite{morris} & $0.0115$ & $0.0085$ \\
    Cephes \cite{cephes}              & $0.0136$ & $0.0119$ \\
    Algorithm 327 \cite{koelbigDilog} & $0.0229$ & $0.0172$ \\
    Chebyshev \cite{luke,root}        & $0.0718$ & $0.0515$ \\
    Algorithm 490 \cite{ginsberg}     & $0.1057$ & $0.0576$ \\
    GSL \cite{gsl}                    & $0.2091$ & $0.0898$ \\
    \bottomrule
  \end{tabular}
  \label{tab:runtime}
\end{table}%

We find that the implementation presented in this paper and the
approach presented in Ref.~\cite{morris} (DILOG 0011) are fastest
among the ones shown in the table.  Both approaches use a rational
function approximant with the same number of terms.  However, the
approximant from Ref.~\cite{morris} has been implemented (for
comparison) to use Horner's scheme to evaluate the numerator and the
denominator.  Thus, the run\-/time difference of $\sim 30\%$ between
the two stems only from the fact that the implementation presented
in this paper uses Estrin's scheme to evaluate the numerator and the
denominator and as such can make use of ILP, while in the
implementation of Ref.~\cite{morris} the numerator and the denominator
must be evaluated sequentially.

The third fastest implementation is provided by Cephes \cite{cephes},
which implements a rational function approximation for Spence's
function $\operatorname{Sp}(x)=\Li(1-x)$.  The implementation provided
by Cephes is by a factor two slower than the implementation prestend
in this paper.  This longer run\-/time has several reasons: For one,
Cephes uses a rational function approximation with more terms in the
numerator and denominator.  Furthermore, Cephes evaluates the
numerator and denominator polynomials using Horner's scheme, which is
purely sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$
  twice, which leads to a further increased run\-/time when a benchmark
  interval with $x>1$ is chosen.}

Algorithm 327 \cite{koelbigDilog}, Algorithm 490 \cite{ginsberg} and
the implementation found in the GNU Scientific Library (GSL)
\cite{gsl} use series expansions, which are summed using Horner's
scheme.  These algorithms thus do not make use of ILP, which is one
reason for their longer run\-/time.  Furthermore, Algorithm 490 and most
series expansions in the GSL use floating\-/point divisions inside
loops, which may be another reason for the increased run\-/time.

A representation in terms of Chebyshev polynomials \cite{luke} is
implemented in the ROOT program library \cite{root}.  In this
implementation the polynomials are summed using Clenshaw's algorithm,
which is also purely sequencial and slightly more complex than
Horner's scheme.  This is a reason why the implementation in
\cite{root} has a larger run\-/time than the rational function
approximants discussed above.

@@

In this paper different approaches for the numerical calculation of
the real dilogarithm have been compared w.r.t.\ their potential for
the use of instruction\-/level (ILP) parallelism.  We found that a
rational minimax approximant in combination with Estrin's scheme to
evaluate the numerator/denominator polynomials provides both a
sufficient numerical precision and a high degree of ILP, potentially
leading to a reduced run\-/time.  An approximant and a corresponding C
implementation have been proposed and compared to several existing
implementations w.r.t.\ their run\-/time.  The C implementation of the
presented approximant can be found in the appendix.  C++ and FORTRAN
implementations can be found in the Polylogarithm library
\cite{polylogarithm}.  Implementations in Julia and Rust can be found
in \cite{PolyLog.jl} and \cite{polylog}, respectively.
